Decoding the Amplitude-Modulated Part the DCF77 Longwave Time Signal by using the Goertzel Algorithm

### 2024-08-16 paper/main.tex
DCF77 signal part finished.

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc,patterns,patterns.meta}
\usepackage{siunitx}  
\usepackage{placeins}
\sisetup{locale = US}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\section{Introduction}
Radio controlled clocks are very popular because they dont require periodical re-adjustment in comparison to common mechanical or quartz-powered clocks.
Instead, the information about the current time is obtained from a RF-signal that is transmitted by an official authority that ensures its availability and
correctness.
The probably most important time signal in Europe is transmitted by the station DCF77 from Mainflingen, a city in Germany that is close to Frankfurt am Main.
The station is operated by a private company that is commissioned by the Physikalisch-Technische Bundesanstalt (PTB) which also supplies the current time information,
generated by two caesium clocks and two caesium fountains. Its operating frequency lies in the long wave area at \SI{77.5}{\kilo\hertz} and the time signal
uses simple modulation techniques (amplitude and phase-modulation) and transmits with \SI{100}{\kilo\watt} Effective Isotropic Radiated Power (EIRP) to enable the reception of the signal in relatively close distances (up to \SI{2000}{\kilo\meter})
with simple technical equipment \cite{b2}.
\begin{figure}[!htbp]
    \centerline{\includegraphics[width=0.3\textwidth]{img/dcf77_range.jpg}}
    \caption{Coverage of the DCF77 signal. Adapted from \cite{b1}}
    \label{fig:upchirp}
\end{figure}
\FloatBarrier
End-user devices usually make use of the amplitude-modulated part of the signal by using an amplitude modulation receiver, often a by employing specialized integrated circuit,
together with a rather big ferrite-rod antenna to obtain the time signal.
This time signal then needs further hardware, often in form of a microcontroller used as the applicaton processor, to decode and retrieve the actual time information.
With the increasing availability of computation power and Digital Signal Processing (DSP) capabilities inside cheap mixed-signal microcontrollers, which are already used
to process the time information, the idea to replace to replace the analog receiver portion with a Software Defined Radio (SDR) is more or less obvious.
Moving the receiver into the application processor would allow for a simpler hardware design by shrinking down the complexity and the parts count of the
external analog circuitry. In fact, only a tuned antenna with a basic pre-amplifier is needed.

@@

\section{The DCF77 signal}
The DCF77 station operates in the long wave area at a frequency of \SI{77.5}{\kilo\hertz}.
Its operation started with transmitting a frequency standard in 1959.
In 1973, the transmission of the current time information was added to the signal by using amplitude modulation.
To allow a reception under more challenging conditions and with a higher resolution and accuracy, the same time information is also added to the signal since 1983 by using phase-modulation \cite{b3}.
The two different parts of the signal that result out of the parallel amplitude and the phase-modulation can be used for different use-cases that are separated by the
requirements regarding the resolution, accuracy, and the reliability of the obtained time information. 
Whereas the amplitude-modulated part is mainly used for general-purpose time application with low resolution requirements, the phase-modulated part is usually used
for specialized applications that require a higher resolution, accuracy, and reliability.
Although being considered as low in comparison to what is achievable by analyzing the phase-modulated part, the usually observed accuracy uncertainity when using the
amplitude-modulated part it is still well below \SI{1}{\second}, being \SI{100}{\milli\second} \cite{b2}.
When using the phase-modulated part, a standard deviation of \SIrange{\pm 2}{\pm 22}{\micro\second} from the atomic clock to the receving device can be observed \cite{b4}.
\par
As the amplitude-modulated part allows simplistic receiver designs through its straightfoward modulation scheme, it is used in the principle presented in this paper
and a focus is put on explaining this part of the DCF77 signal.

\section{Principle of a DCF77 receiver}
\begin{figure}[!h]
    \centering
    \resizebox{0.48\textwidth}{!}{
        \begin{tikzpicture}
            \clip (-11.5,-1) rectangle + (23,5); %Clipping: where to start, shape and size

@@

                %DCF77 at -f_c
                \draw[thick]
                ($-1*(fc) + (-0.12,0) + \x*(fs)$) -- % Left line start
                ++(0,1.5) -- %Left line up
                ++(0.24,-0.5) -- %Roof
                ++(0,-1) -- %Right line down
                cycle; % Close the shape
                \draw[very thick] ($-1*(fc) + (0, -0.2) + \x*(fs)$) -- ++(0, 0.4); %Line x-axis
                \node at ($-1*(fc) + (0, -0.8) + \x*(fs)$) {$-f_{c} \pgfmathprintnumber[print sign]{\x} f_{s}$}; %Index x-axis
            };
        \end{tikzpicture}
    } 
\end{figure}

@@

\begin{figure}[htbp]
\centerline{\includegraphics{img/fig1.png}}
\caption{Example of a figure caption.}
\label{fig}
\end{figure}

@@

\begin{thebibliography}{00}
\bibitem{b1} -, ``DCF77,'' 2023, Wikipedia entry. [Online]. Available: https://de.wikipedia.org/wiki/DCF77
\bibitem{b2} D. Priester and P. Hetzel and A. Bauch, ``Zeit- und Normalfrequenzverbreitung mit DCF77,'' PTB Mitteilungen, 2004, [Online]. Available: https://www.ptb.de/cms/fileadmin/internet/fachabteilungen/abteilung\_4/4.4\_zeit\_und\_frequenz/4.42/dcf77.pdf
\bibitem{b3} D. Priester and P. Hetzel and A. Bauch, ``Zeit- und Normalfrequenzverbreitung mit DCF77: 1959-2009 und darüber hinaus,'' PTB Mitteilungen, 2009, [Online]. Available: https://www.ptb.de/cms/fileadmin/internet/fachabteilungen/abteilung\_4/4.4\_zeit\_und\_frequenz/pdf/2009\_Bauch\_PTBM\_\_DCF77.pdf
\bibitem{b4} P. Hetzel, ``Time dissemination via the LF transmitter DCF77 using a pseudo-random phase-shift keying of the carrier,'' PTB Braunschweig, 1988, [Online]. Available: https://www.ptb.de/cms/fileadmin/internet/fachabteilungen/abteilung\_4/4.4\_zeit\_und\_frequenz/pdf/5\_1988\_Hetzel\_-\_Proc\_EFTF\_88.pdf

%\bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350.
%\bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished.
%\bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press.
%\bibitem{b6} Y. Yorozu, M. Hirano, K. Oka, and Y. Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl. J. Magn. Japan, vol. 2, pp. 740--741, August 1987 [Digests 9th Annual Conf. Magnetics Japan, p. 301, 1982].
%\bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
%\bibitem{b9} S. Liu, ``Wi-Fi Energy Detection Testbed (12MTC),'' 2023, gitHub repository. [Online]. Available: https://github.com/liustone99/Wi-Fi-Energy-Detection-Testbed-12MTC
%\bibitem{b10} ``Treatment episode data set: discharges (TEDS-D): concatenated, 2006 to 2009.'' U.S. Department of Health and Human Services, Substance Abuse and Mental Health Services Administration, Office of Applied Studies, August, 2013, DOI:10.3886/ICPSR30122.v2
%\bibitem{b11} K. Eves and J. Valasek, ``Adaptive control for singularly perturbed systems examples,'' Code Ocean, Aug. 2023. [Online]. Available: https://codeocean.com/capsule/4989235/tree
\end{thebibliography}

>>>>

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{placeins}

\usepackage{tikz}
\usetikzlibrary{calc,patterns,patterns.meta}

\usepackage{siunitx} 
\sisetup{locale = US}
\DeclareSIUnit{\Bit}{Bit}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

    \begin{document}

@@

\section{Introduction}
Radio controlled clocks are very popular because they dont require periodical re-adjustment in comparison to common mechanical or quartz-powered clocks.
Instead, the information about the current time is obtained from a RF-signal that is transmitted by an official authority that ensures its availability and
correctness.
The probably most important time signal in Europe is transmitted by the station DCF77 from Mainflingen, a city in Germany that is close to Frankfurt am Main.
The station is operated by a private company that is commissioned by the Physikalisch-Technische Bundesanstalt (PTB) which also supplies the current time information,
generated by two caesium clocks and two caesium fountains. Its operating frequency lies in the long wave area at \SI{77.5}{\kilo\hertz} and the time signal
uses simple modulation techniques (amplitude and phase-modulation) and transmits with \SI{100}{\kilo\watt} Effective Isotropic Radiated Power (EIRP) to enable the reception of the signal in relatively close distances (up to \SI{2000}{\kilo\meter})
with simple technical equipment \cite{b2}.
\begin{figure}[!htbp]
    \centerline{\includegraphics[width=0.3\textwidth]{img/dcf77_range.jpg}}
    \caption{Coverage of the DCF77 signal. Adapted from \cite{b1}}
    \label{fig:dcf77_range}
\end{figure}
\FloatBarrier
End-user devices usually make use of the amplitude-modulated part of the signal by using an amplitude modulation receiver, often a by employing specialized integrated circuit,
together with a rather big ferrite-rod antenna to obtain the time signal.
This time signal then needs further hardware, often in form of a microcontroller used as the applicaton processor, to decode and retrieve the actual time information.
With the increasing availability of computation power and Digital Signal Processing (DSP) capabilities inside cheap mixed-signal microcontrollers, which are already used
to process the time information, the idea to replace to replace the analog receiver portion with a Software Defined Radio (SDR) is more or less obvious.
Moving the receiver into the application processor would allow for a simpler hardware design by shrinking down the complexity and the parts count of the
external analog circuitry. In fact, only a tuned antenna with a basic pre-amplifier is needed.

@@

\section{The DCF77 signal}
The DCF77 station operates in the long wave area at a frequency of \SI{77.5}{\kilo\hertz}.
Its operation started with transmitting a frequency standard in 1959.
In 1973, the transmission of the current time information was added to the signal by using amplitude modulation.
To allow a reception under more challenging conditions and with a higher resolution and accuracy, the same time information is also added to the signal since 1983 by using phase-modulation \cite{b3}.
The two different parts of the signal that result out of the parallel amplitude and the phase-modulation can be used for different use-cases that are separated by the
requirements regarding the resolution, accuracy, and the reliability of the obtained time information. 
Whereas the amplitude-modulated part is mainly used for general-purpose time application with low resolution requirements, the phase-modulated part is usually used
for specialized applications that require a higher resolution, accuracy, and reliability.
Although being considered as low in comparison to what is achievable by analyzing the phase-modulated part, the usually observed accuracy uncertainity when using the
amplitude-modulated part it is still well below \SI{1}{\second}, being \SI{100}{\milli\second} \cite{b2}.
When using the phase-modulated part, a standard deviation of \SIrange{\pm 2}{\pm 22}{\micro\second} from the atomic clock to the receving device can be observed \cite{b4}.
\par
As the amplitude-modulated part allows simplistic receiver designs through its straightfoward modulation scheme, it is used in the principle presented in this paper
and therefore a focus is put on explaining this part of the DCF77 signal.
The time information, represented as a series of binary symbols, is modulated onto the carrier with a frequency of \SI{77.5}{\kilo\hertz} by using
Amplitude-Shift Keying (ASK). Every symbol has the duration of exactly \SI{1}{\second}, with a part that is either \SI{100}{\milli\second} or \SI{200}{\milli\second}
long where it has a value to zero.
This the duration of this "gap" is used to encode respectively a 0 or a 1 and effectively switches the strength of the carrier between \SI{15}{\percent} and \SI{100}{\percent} \cite{b5}.
\begin{figure}[htbp]
    \centerline{\includegraphics[width=0.45\textwidth]{img/dcf77_symbol_encoding.png}}
    \caption{Symbol encoding of the DCF77 signal. From \cite{b1}}
    \label{fig:dcf77_symbol_encoding}
\end{figure}
\FloatBarrier\noindent
To allow the detection of a new minute, the gap is left out between the last (59) second of the ongoing and the first (0) second of the next minute.
By using this scheme, a dataframe consisting out \SI{59}{\Bit}, starting at second one, can be transmitted every minute.
Bit \SIrange{0}{19}{} are used to send encrypted weather data, to notify the PTB in case of errors in the transmitter, to announce an upcoming
switch to or from Daylight Saving Time (DST), to show if DST is currently enabled or not, and to announce an upcoming leap second \cite{b5}.
As the decryption of the weather date requires a specialized third-party integrated circuit and the notification bit is only important for the PTB, these parts
of the frame are discarded by most applications.
The leftover Bits \SIrange{20}{58}{} are used to represent the current time and date in Binary Coded Decimal (BCD) format, including parity bits for recognizing errors.
\begin{figure}[htbp]
    \centerline{\includegraphics[width=0.35\textwidth]{img/dcf77_frame_encoding.jpg}}
    \caption{Data frame encoding of the DCF77 signal. From \cite{b5}}
    \label{fig:dcf77_data_encoding}
\end{figure}
\FloatBarrier\noindent
This means that a receiver needs to run at least for \SI{39}{\second} after powerup to receive one whole block containing the current time information.
In reality, this period is usually longer because the receiver needs to adapt to the current receiving conditions. 
These conditions underly a lot of influences, including temporary fading, electromagnetic noise from the close environment or varying signal strengths
during different times of the day.

\section{Principle of a DCF77 receiver}
\begin{figure*}[!htbp]
    \centering
    \resizebox{0.9\textwidth}{!}{
        \begin{tikzpicture}
            \clip (-11.5,-1) rectangle + (23,5); %Clipping: where to start, shape and size

@@

                %DCF77 at -f_c
                \draw[thick]
                ($-1*(fc) + (-0.12,0) + \x*(fs)$) -- % Left line start
                ++(0,1.5) -- %Left line up
                ++(0.24,-0.5) -- %Roof
                ++(0,-1) -- %Right line down
                cycle; % Close the shape
                \draw[very thick] ($-1*(fc) + (0, -0.2) + \x*(fs)$) -- ++(0, 0.4); %Line x-axis
                \node at ($-1*(fc) + (0, -0.8) + \x*(fs)$) {$-f_{c} \pgfmathprintnumber[print sign]{\x} f_{s}$}; %Index x-axis
            };
        \end{tikzpicture}
    } 
\end{figure*}

@@

@@

\begin{thebibliography}{00}
\bibitem{b1} -, ``DCF77,'' 2024, Wikipedia entry. [Online]. Available: https://de.wikipedia.org/wiki/DCF77
\bibitem{b2} D. Priester and P. Hetzel and A. Bauch, ``Zeit- und Normalfrequenzverbreitung mit DCF77,'' PTB Mitteilungen, 2004, [Online]. Available: https://www.ptb.de/cms/fileadmin/internet/fachabteilungen/abteilung\_4/4.4\_zeit\_und\_frequenz/4.42/dcf77.pdf
\bibitem{b3} D. Priester and P. Hetzel and A. Bauch, ``Zeit- und Normalfrequenzverbreitung mit DCF77: 1959-2009 und darüber hinaus,'' PTB Mitteilungen, 2009, [Online]. Available: https://www.ptb.de/cms/fileadmin/internet/fachabteilungen/abteilung\_4/4.4\_zeit\_und\_frequenz/pdf/2009\_Bauch\_PTBM\_\_DCF77.pdf
\bibitem{b4} P. Hetzel, ``Time dissemination via the LF transmitter DCF77 using a pseudo-random phase-shift keying of the carrier,'' PTB Braunschweig, 1988, [Online]. Available: https://www.ptb.de/cms/fileadmin/internet/fachabteilungen/abteilung\_4/4.4\_zeit\_und\_frequenz/pdf/5\_1988\_Hetzel\_-\_Proc\_EFTF\_88.pdf
\bibitem{b5} -, ``DCF77 webpage,'' 2024, Official page of the PTB [Online]. Available: https://www.ptb.de/cms/en/ptb/fachabteilungen/abt4/fb-44/ag-442/dissemination-of-legal-time/dcf77.html
%\bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350.
%\bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished.
%\bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press.
%\bibitem{b6} Y. Yorozu, M. Hirano, K. Oka, and Y. Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl. J. Magn. Japan, vol. 2, pp. 740--741, August 1987 [Digests 9th Annual Conf. Magnetics Japan, p. 301, 1982].
%\bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
%\bibitem{b9} S. Liu, ``Wi-Fi Energy Detection Testbed (12MTC),'' 2023, gitHub repository. [Online]. Available: https://github.com/liustone99/Wi-Fi-Energy-Detection-Testbed-12MTC
%\bibitem{b10} ``Treatment episode data set: discharges (TEDS-D): concatenated, 2006 to 2009.'' U.S. Department of Health and Human Services, Substance Abuse and Mental Health Services Administration, Office of Applied Studies, August, 2013, DOI:10.3886/ICPSR30122.v2
%\bibitem{b11} K. Eves and J. Valasek, ``Adaptive control for singularly perturbed systems examples,'' Code Ocean, Aug. 2023. [Online]. Available: https://codeocean.com/capsule/4989235/tree
\end{thebibliography}
